---
title: "R Notebook - Descibing Data II"
output: html_notebook
---

# One Quantitative Variable

## Histograms

Graphs for 1 Quantitative var

```{r - Histograms}
#fetching the dataset
data(airquality)
#Compactly display the internal structure of the object
str(airquality)

```

Choosing a "feature" a.k.a Quantitative var = Temperature

```{r}
Temperature <- airquality$Temp
hist(Temperature)

```

If we want to have the probability distribution instead of the actual frequency, we can set the parameter 'freq=F'

```{r}
hist(Temperature, freq = F)
```

The hist() function returns a list of 6 objects :

-   \$breaks: Bins of the graph

-   \$counts: Frequency (for each bin?- to check)

-   \$density: probability

-   \$mids (mid points of the cells)

-   xname: x arguments of the cells

-   equidist: Bollean indicating that the breaks are equally spaced

We can use the return values of the hist for further processing

```{r}
#use the return from hist to complete the histogram
h = hist(Temperature, ylim = c(0,40))
#see the contents of h (previous hist)
str(h)
# use these to complete the graph
text(h$mids, h$counts, labels = h$counts, adj=c(0.5, -0.5))

```

#### Truehist() a Hist() alternative

Scales the counts to give an estimate of the probability

```{r}
th = truehist(Temperature)
th# th doesn't return any value
```

## Shape

Measures of central tendency

-   Mean

-   Median - Middle of a distribution that is ordered from lowest to highest value in the X-axis

-   Mode - Value with the highest frequency

Note:R does not have a standard in-built function to calculate mode. So we create a user function to calculate mode of a data set in R. This function takes the vector as input and gives the mode value as output.

Simple example

```{r}
ex = c(5,7,8,9,9,11,13)
# function to calculate the mode
getmode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}
ex_ct=c(mean(ex), median(ex), getmode(ex))
str(ex_ct)
```

## Spread

Measures of Spread

-   Variance

-   Standard deviation (most used where we are handling quantitative data and an aprox. normal distribution)

Standard deviation = SQRT(Variance)

### Standard Deviation

Roughly the avg. difference between the values and the mean

-   denoted by an "s" or a greek "rhó" when referring to the population

How it's calculated:

1.  Compute the sample mean
2.  Subtract the sample mean from each individual value
3.  square (\^2) each deviation (result from 2)
4.  Sum all deviations
5.  divide the sum of squares by n-1 (= Sample variance)
6.  SQRT the result

```{r}
data=c(5,7,8,9,9,11,13) 
data
# Step 1: compute the mean 
x.mean=mean(data)
x.mean
# Step 2: compute the deviations 
x.deviations=data-x.mean 
x.deviations
# Step 3: square de deviations 
x.deviations.squared=x.deviations^2
x.deviations.squared
# Step 4: sum the squared deviations 
x.SS = sum(x.deviations.squared) 
x.SS
# Step 5: Divide by n-1 to compute variance 
x.variance=x.SS/6
x.variance
# Step 6: Take the square root of the variance 
x.s=sqrt(x.variance)
x.s
```

### Empirical Rule

It's a statement applicable in **normal distributions**

So in a **normal distribution** we can assume:

1.  Aprox. 68% of the data will be within 1 Std. deviation of the mean
2.  Aprox. 95% of observations fall within 2 Std. deviations of the mean
3.  Aprox. 99,7% will be within 3 std deviations of the mean

```{r} ### to include an example in R of a normal distribution}

```

## Z-Cores

Z-Score describes one observation in relation to the distribution of all observations.

It can also be seen as the distance of an observation to the mean in std. deviations -\> a.k.a. Standardized score

##### Remarks

-   The Z-Score this allows an abstraction to the actual units of measure of the variables we are analyzing

-   Usually relevant to analyze individuals or sub-populations of the distribution

### z-distribution or Standard Normal Distribution

A bell-shaped distribution with a mean of 0 and standard deviation of 1, also known as the standard normal distribution

## Percentiles

We can determine a proportion of values falling below a given value, e.g. a student scores better than 90% then the student falls on the 90th percentile.

## Five Number Summary

Minimum, Q1, Median, Q3 Maximum

-   Q1 = 25th percentile or 1st quartile

-   Q2 = 75th percentile or 3rd quartile

These are used to describe some key figures of a distribution. Using the 5 number summary we can also calculate the range and interquartile range

#### Remark

If we we are given, or able to obtain, the 5 numbers, we have basically everything about a distribution.

```{r}

```
