---
title: "SfEDA selected exercises - Lab 3"
author: "Jorge M. Mendes and M. Helena Baptista"
date: "2021"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
library("readxl")
library(plyr)
library(knitr)
library(exams)
```

# Exercise 1

The *INTERNET* data file contains data from 2010 relating to gross
domestic product (GDP) per capita in US\$ thousands and the percentage
of the population that are Internet users for 212 countries. Here, GDP
is based on purchasing power parities to account for between-country
differences in price levels. This problem investigates whether there is
a linear association between these two variables. In particular, how
effective is it to use to predict using simple linear regression?

1.  Find the least squares line for the data, that is, use *R* to find
    the intercept and slope of the least squares line, and write down
    the equation of that line.

```{r echo = T, comment = NA}
library(readr)
internet <- read_csv("datasets/internet.csv")
View(internet)
```

```{r echo = T, comment = N}
model1 <- lm(Int ~ Gdp, data = internet)
summary(model1)
plot(model1$fitted.values, model1$residuals)
plot(model1$residuals, internet$Int)
hist(model1$residuals)
qqnorm(model1$residuals, main="Normal Q-Q Plot of Std Residuals")
qqline(model1$residuals)
```

2.  Interpret the estimates of the slope and the intercept in the
    context of the problem.

    Slope: We have a slope value of 1,36

    Intercept: 12,36 (note that it doesn't make sense to consider a GDP
    of 0 in any cases)

    Regression line equation = 12,36 + 1,36 \* Xi

Model Quality:

-   Std. Error: 18.14 (We can expect that our predicted values can vary
    ± multiplier \\\**18.14 - for a 95% conf. interval 1,96\\\*18.14*)

-   Coefficient Determination - R- Squared: 0.58 (GDP can explain 58% of
    the variability of the internet usage)

-   Correlation:

    -   our p value is \< 0.05 so we can reject the null hypothesis - "
        there is a non 0 slope"

2.  Predict the percentage of Internet users if GDP per capita is
    US\$20,000.

```{r echo = T, comment = N}
Country <- c("Country x")
Gdp <- c(20)
new_data <- data.frame(Country, Gdp)
predict(model1, newdata = new_data)

```

4.  Draw a scatterplot with Internet on the vertical axis and GDP on the
    horizontal axis, and add the least squares line to the plot.

```{r echo = T, comment = N}
library(ggplot2)
library(ggpubr)
ggplot(data = internet, aes(x = Gdp, y = Int)) + 
  geom_smooth(method = "lm", se=FALSE, color="black", formula = y ~ x) +
        geom_point()+
        stat_regline_equation(label.y = 110) #this means at 110th unit regresion line equation will be shown
```

5.  Based on the scatterplot, do you think it is appropriate to use this
    simple linear regression model in this problem, or is the model
    potentially misleading (and if so, how)?

Our model and best-fit line may be misleading, considering the graph and
the 58% R2 value.

# Exercise 2

The *MLB* data file contains the number of wins and the batting average
calculated as number of hits divided by number of at-bats) for the 2010
regular season. This problem is a highly simplified introduction to
which uses baseball statistics to find objective insights into the game.

1.  Remove the American League teams from the data file to leave just
    the 16 teams in the National League. Consider using simple linear
    regression to predict the number of wins from a team's batting
    average. Would you expect the slope of the resulting least squares
    line to be positive or negative? Explain.

```{r echo = T, comment = N}
library(readr)
mlb <- read_csv("datasets/mlb.csv")
View(mlb)
```

```{r echo = T, comment = N}
unique(mlb$League)
library(dplyr)
national_league <- mlb %>% filter(League == "NL")
cor(national_league$Win, national_league$Bat)

```

```{r}
model2 <- lm(Bat ~ Win, data = national_league)
summary(model1)
plot(model2$fitted.values, model2$residuals)
plot(model2$residuals, national_league$Win)
hist(model2$residuals)
qqnorm(model2$residuals, main="Normal Q-Q Plot of Std Residuals")
qqline(model2$residuals)
```

2.  Construct an appropriate scatterplot of the data. Does the pattern
    of points in the scatterplot agree with your answer from part 1.?

```{r echo = T, comment = N}

ggplot(national_league, aes(x=Bat, y=Win)) + geom_point()
```

3.  Find the intercept and slope of the least squares line and write
    down the equation of that line.

```{r echo = T, comment = N}
model2 <- lm(Win~Bat, data = national_league)
summary(model2)

```

4.  Add the least squares line to the scatterplot from part 2.. Does the
    least squares line seem to adequately represent the dominant pattern
    in the points on the scatterplot?

```{r echo = T, comment = N}
ggplot(national_league, aes(x=Bat, y=Win)) + geom_point() +
  geom_smooth(method = "lm", se=FALSE, color="black", formula = y ~ x) +
        geom_point()+
        stat_regline_equation(label.y = 110) #this means at 110th unit regresion line equation will be shown
```

5.  Does the number of wins appear to be strongly associated with a
    team's batting average? Explain.

```{r echo = T, comment = N}
confint(model2)
```

6.  Interpret the estimates of the slope and the intercept in the
    context of the problem.

7.  Reopen the data file, remove the National League teams from the data
    file to leave just the 14 teams in the American League, and repeat
    parts 1. to 6.

# Exercise 3

The *ELECTRICITY* data file contains data on electricity consumption in
billions of kilowatt-hours (Elec) and gross domestic product (GDP) in
billions of dollars for the 30 most populous countries. Here, GDP is
based on purchasing power parities to account for between-country
differences in price levels. The data file can be used to investigate
the claim that there is a linear association between electricity
consumption and GDP. For the purposes of this problem, assume that
increases in electricity consumption tend to respond to increases in GDP
(rather than the other way around).

1.  Say which variable should be the predictor variable ($X$) and which
    the response variable ($Y$) for this problem, and consider the
    simple linear regression model $E(Y)=b_0+b_1X$.

\*Before looking at the data, say whether the population value would be
positive or negative under the claim that electricity consumption tends
to increase in response to increases in GDP.

2.  Plot the data in a scatterplot (make sure that you put the
    appropriate variables on each axis). Add the least squares line to
    the scatterplot. Briefly describe the dominant pattern in the points
    on the scatterplot and their position in relation to the least
    squares line.

```{r echo = T, comment = N}
library(readr)
electricity <- read_csv("electricity.csv")
View(electricity)

ggplot(electricity, aes(x=Gdp, y=Elec)) + geom_point() +
  geom_smooth(method = "lm", se=FALSE, color="black", formula = y ~ x) +
        geom_point()+
        stat_regline_equation(label.y = 2000) #this means at 110th unit regresion line equation will be shown

```

3.  Identify the countries with the two highest values of *Gdp*, remove
    them from the dataset, and redraw the scatterplot and least squares
    line. How does your visual impression of the scatterplot change?

*Hint: Focus more on what you see overall, not so much on individual
points. In particular, think about whether the overall trend changes and
whether the variation of the points about the line changes.*

```{r echo = T, comment = N}
ggplot(electricity, aes(x=Gdp, y=Elec)) + geom_point() +
geom_text(
    label=electricity$Country
  )

reduced <- electricity %>% filter(Elec < 1000)

ggplot(reduced, aes(x=Gdp, y=Elec)) + geom_point() +
  geom_smooth(method = "lm", se=FALSE, color="black", formula = y ~ x) +
        geom_point()+
        stat_regline_equation(label.y = 750) #this means at 110th unit regresion line equation will be shown
```

4.  Fit the appropriate simple linear regression model to the dataset
    that has the countries with the two highest values of *Gdp* removed.
    Do the results provide evidence that the claim of a positive
    association could be true for the remaining 28 countries? *Hint:
    This problem is asking you to do a hypothesis test. Use a
    significance level of 5%, make sure you write down the null and
    alternative hypotheses, and conclude with a statement either
    supporting or contradicting the claim once you have written up the
    results of the hypothesis test.*

5.  Based on the graph from part 3. and the model from part 4., which
    country has a particularly *hight* electricity consumption and which
    country has a particularly *low* consumption *relative to the
    dominant pattern for the remaining countries*?

# Exercise 4

In the classes you have been told three numerical measures for
evaluating a simple linear regression model: the regression standard
error, $s$, the coefficient of determination, $R^2$, and the
$t$-statistic for the slope parameter, $b_1$. The purpose of this
problem is to demonstrate that a simple linear regression model with
superior measures of fit (i.e., a lower $s$, a higher $R^2$, and a
higher absolute $t$-statistic,$|t_{b_1}|$) does not necessarily imply
that the model is more appropriate than a model with a higher a lower
$R^2$, and a lower absolute $t$-statistic, $|t_{b_1}|$. These three
measures of model fit should always be used in conjunction with a
graphical check of the assumptions to make sure that the model is
appropriate. Use the simulated data from the COMPARE data file, which
contains a single response variable, and four possible predictor
variables, Xi, X3, and X4.

1.  Fit a simple linear regression using $X1$ as the predictor. Make a
    note of the values of the regression standard error, $s$, the
    coefficient of determination, $R^2$, and the $t$-statistic for the
    slope parameter, $t_{b_1}$. Also construct a scatterplot of
    (vertical) versus $X_1$ (horizontal) and add the least squares
    regression line to the plot.

2.  Repeat part 1., but this time use $X_2$ as the predictor. Do the
    values of $R^2$, and $t_{b_1}$ suggest a worse or better fit than
    the model from part 1.? Does the visual appearance of the
    scatterplot for the $X_2$ model confirm or contradict the numerical
    findings?

3.  Repeat part 3., but this time use $X_3$ as the predictor. Do the
    values of $R^2$, and $t_{b_1}$ suggest a worse or better fit than
    the model from part 1.? Does the visual appearance of the
    scatterplot for the $X_3$ model confirm or contradict the numerical
    findings?

4.  Repeat part 4., but this time use $X_4$ as the predictor. Do the
    values of $R^2$, and $t_{b_1}$ suggest a worse or better fit than
    the model from part 1.? Does the visual appearance of the
    scatterplot for the $X_4$ model confirm or contradict the numerical
    findings?

# Exercise 5 (2.10:IP)

A restaurant owner wishes to model the association between mean daily
costs, $\text{E}(Costs)$ , and the number of customers served, $Covers$,
using the simple linear regression model,
$\text{E}(Costs)=b_0+b_1Covers$. The owner collected a week of data and
the values for $n = 7$ days are shown in the accompanying table,
followed by computer output for the resulting regression analysis.

```{r}
Costs=c(1000,2180,2240,2410,2590,2820,3060)
Covers=c(0,60,120,133,143,175,175)
dt=data.frame(Costs=Costs, Covers=Covers)
kable(dt)
#
m1 = lm(Costs~Covers, data=dt)
summary(m1)
```

1.  State the set of hypotheses that you would test to determine whether
    costs are positively linearly associated with the number of
    customers.
2.  Is there sufficient evidence of a positive linear association
    between costs and number of customers? Use significance level 5%.
3.  Give a practical interpretation of the estimate of the slope of the
    least squares line.
4.  Give a practical interpretation, if you can, of the estimate of the
    intercept of the least squares line.
5.  Give a practical interpretation of $s$, the estimate of the standard
    deviation of the random error term in the model.
6.  Give a practical interpretation of $R^2$, the coefficient of
    determination for the simple linear regression model.

# Exercise 6 (3.1:IP)

The *MOVIES* data file contains data on 25 movies from "The Internet
Movie Database" (www.imdb.com). Based on this dataset, we wish to
investigate whether alltime U.S. box office receipts (*Box* in millions
of U.S. dollars unadjusted for inflation) are associated with any of the
following variables:

*Rate* = Internet Movie Database user rating (out of 10)

*User* = Internet Movie Database users rating the movie (in thousands)

*Meta* = "metascore" based on 35 critic reviews (out of 100)

*Len* = runtime (in minutes)

*Win* = award wins

*Nom* = award nominations

Theatrical box office receipts (movie ticket sales) may include
theatrical re-release receipts, but exclude video rentals, television
rights, and other revenues.

1.  Write out an equation for a multiple linear regression model for
    predicting response from just three predictors: *Rate*, *User*, and
    *Meta*. *Hint: This question is asking you to write an equation for*
    $\text{E}(Box)$.
2.  Use *R* to fit this model and write out the estimated multiple
    linear regression equation [i.e., replace the b's in part 1. with
    numbers].
3.  Interpret the estimated regression parameter for in the context of
    the problem [i.e., put the appropriate number from part 2. into a
    meaningful sentence, remembering to include the correct units for
    any variables that you use in your answer].

# Exercise 7 (3.2:IP)

Consider the **MOVIES** data file from Problem 6 again.

1.  Use *R* to fit the following (complete) model for as a function of
    all six predictor variables:
    $$\text{E}(Box)=b_0+b_1Rate+b_2+User+b_3Meta+b_4Len+b_5Win+b_6Nom.$$
    Write down the residual sum of squares for this model.
2.  Use *R* to fit the following (reduced) model:
    $$\text{E}(Box)=b_0+b_1Rate+b_2+User+b_3Meta.$$ [This is the model
    from Problem 6 part 2.] Write down the residual sum of squares for
    this model.
3.  Using the results from parts 1. and 2. together with the nested
    model test $F$-statistic formula, test the null hypothesis
    $H_0: b_4=b_5=b_6= 0$ in the complete model, using significance
    level 5%. Write out all the hypothesis test steps and interpret the
    result in the context of the problem.
4.  Check your answer for part 3. by using *R* to do the nested model
    test directly. State the values of the $F$-statistic and the
    $p$-value, and draw an appropriate conclusion.
5.  Another way to see whether we should prefer the reduced model for
    this example is to see whether the regression standard error is
    smaller for the reduced model than for the complete model and
    whether adjusted $R^2$ is higher for the reduced model than for the
    complete model. Confirm whether these relationships hold in this
    example (i.e., compare the values of $s$ and adjusted $R^2$ in the
    reduced and complete models).

# Exercise 8 (3.2:IP)

Researchers at General Motors analyzed data on 56 U.S. Standard
Metropolitan Statistical Areas (SMSAs) to study whether air pollution
contributes to mortality. These data are available in the **SMSA** data
file and were obtained from the "Data and Story Library" at
lib.stat.emu.edu/DASL/ (the original data source is the U.S. Department
of Labor Statistics). The response variable for analysis is *Mort* = age
adjusted mortality per 100,000 population (a mortality rate
statistically modified to eliminate the effect of different age
distributions in different population groups). The dataset includes
predictor variables measuring demographic characteristics of the cities,
climate characteristics, and concentrations of the air pollutant nitrous
oxide (NO$_x$).

1.  Fit the (complete) model
    $\text{E}(Mort) = b_0 + b_1Edu + b_2 Nwt + b_3Jant + b_4Rain + b_5Nox+b6Hum+b_7Inc$,
    where $Edu$ is median years of education, *Nwt* is percentage
    nonwhite, *Jant* is mean January temperature in degrees Fahrenheit,
    *Rain* is annual rainfall in inches, *Nox* is the natural logarithm
    of nitrous oxide concentration in parts per billion, *Hum* is
    relative humidity, and *Inc* is median income in thousands of
    dollars. Report the least squares (regression) equation.
2.  Do a nested model $F$-test (using a significance level of 5%) to see
    whether and *Inc* provide significant information about the
    response, *Mort*, beyond the information provided by the other
    predictor variables.
3.  Do individual $t$-tests (using a significance level of 5%) for each
    predictor in the (reduced) model
    $\text{E}(Mort) = b_0 + b_1Edu + b_2 Nwt + b_3Jant + b_4Rain + b_5Nox$.
4.  Check the four model assumptions for the model from part 3. *Hint:
    Hint: Ideally, you should look at six residual plots (five with each
    of the five predictors on the horizontal axis in turn and one with
    the predicted values on the horizontal axis) to check the zero mean,
    constant variance, and independence assumptions. You should also use
    a histogram and/or QQ-plot to check the normality assumption. The
    more comprehensive you can be in checking the assumptions, the more
    confident you can be about the validity of your model.*
5.  Write out the least squares equation for the model from part 3. Do
    the signs of the estimated regression parameters make sense in this
    context?
