---
title: "LEc 20210421"
output: html_notebook
---

# **Exercise 1 - Statistical inference I**

For each question, indicate whether it is best assessed by using a confidence interval, a hypothesis test, or whether statistical inference is not relevant to answer it. If inference is not relevant, explain why.

1.  What percent of US voters support instituting a national kindergarten through 12th grade math curriculum?

2.  Do basketball players hit a higher proportion of free throws when they are playing at home than when they are playing away?

    **AW:** Hypothesis testing.

3.  Do a majority of adults riding a bicycle wear a helmet?

    **AW:**

    -   Confidence interval.

    -   Hypothesis test:

        -   Null Hypothesis: 50% use helmet

        -   Alternative hypothesis: More that 50 % uses helmet.

4.  On average, were the 23 players on the 2010 Canadian Olympic hockey team older than the 23 players on the 2010 US Olympic hockey team?

    **AW:**

    -   No inference. We have all the data.

5.  What proportion of people using a public restroom wash their hands after going to the bathroom?

    **AW:**

    -   Confidence interval

6.  On average, how much more do adults who played sports in high school exercise than adults who did not play sports in high school?

    **AW:**

    -   Hypothesis test.

    -   Confidence interval if you want to know how much more.

7.  In 2010, what percent of the US Senate voted to confirm Elena Kagan as a member of the Supreme Court?

    **AW:**

    -   We have all the data. Proportion.

8.  What is the average daily calorie intake of 20-year-old males?

    **AW:**

    -   Confidence Interval

```{r}
plot(cars)
```

# **Exercise 3 - Effect of Smoking on Pregnancy Rate**

Studies have concluded that smoking while pregnant can have negative consequences, but could smoking also negatively affect one's ability to become pregnant? A study collected data on 678 women who had gone off birth control with the intention of becoming pregnant. Smokers were defined as those who smoked at least one cigarette a day prior to pregnancy. We are interested in the pregnancy rate during the first cycle off birth control. The results are summarized in the following table:

    ##              Smoker     Non-smoker Sum       
    ## Pregnant           38.0      206.0      244.0
    ## Not pregnant       97.0      337.0      434.0
    ## Sum               135.0      543.0      678.0

We wish to estimate the difference in the proportion who successfully get pregnant, between smokers and non-smokers.

1.  Find the best point estimate for the difference in proportions.

    **AW:**

    NOTE: As long as we have a big enough sample the best estimate is the difference in sample proportions

    ```{r}
    p_s <- 38/135
    p_ns <- 206/543
    result <- p_s - p_ns
    result
    ```

2.  Use *R* or other technology to find and interpret a 90% confidence interval for the difference in proportions. Is it plausible that smoking has no effect on pregnancy rate?

    NOTE: as we have 10 observations of each cases: pregnant smokers and non-smokers

    ```{r}
    prop.test(c(38,206), c(135,543), conf.level = 0.9)

    ```

    **AW:**

    Yes, P-value is 0.04 which is lower than our alpha (1-0.9) so we reject our Null hypothesis which is: There's no effect of smoking in the possibility to get pregnant.

Option 2: Bootstrapping:

-   As we only have a contingency table

    ```{r}
    #smoker size
    n_s = 135
    #non-smoker size
    n_ns = 543

    s <- matrix(0, nrow=1, ncol=10000)
    for (i in 1:10000) {
      s[,i] = sum(rbinom(n_s,1,p_s))/n_s - sum(rbinom(n_ns, 1 , p_ns))/n_ns
    }

    hist(s)
    ```

# **Exercise 4 - Standard Error for Proportion of Hollywood Movies that are Action Movies**

Previously the dataset *HollywoodMovies2011*, which contains information on all the 136 movies that came out of Hollywood in 2011. Thirty-two of the Hollywood movies that year were classified as action movies.

1.  What proportion of Hollywood movies in 2011 were action movies? Use the correct notation with your answer.

    **AW:**

    -   

2.  Use *R* to generate a sampling distribution for the proportion of action movies for sample proportions of size n=30n=30. Give the shape and center of the sampling distribution and give the standard error.
